Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-10 intro.tex
Liz is now done with intro

Complicated time-series data are ubiquitous in modern scientific
research.  The complexity of these data spans a wide range.  On the
low end of this spectrum are time series that exhibit perfect
predictive structure, i.e, signals whose future values can be
successfully predicted from past values.  In situations like this,
there is an underlying process that generates information and/or
transmits it from the past to the future in a perfectly predictable
fashion.  Constant or periodic signals, for example, fall in this
class.  On the opposite end of this spectrum are signals that are what
we call \emph{fully complex}, where the underlying generating process
transmits no information at all from the past to the future.  Some
examples of this are white noise or random-walk processes.  In signals
like this, knowledge of the past gives no insight into the future,
\emph{regardless} of what model one chooses to use.  Signals in the
midrange of this spectrum---e.g., deterministic chaos, high-period
signals [[Liz question: why does the rank of the periodicity force
    them off the low end of the spectrum?  Any periodic signal can be
    decomposed into sine waves, which can be predicted perfectly...]]
with some additive noise---pose interesting challenges from a modeling
perspective.  In these signals, enough information is being
transmitted from the past to the future that an \emph{ideal}
model---one that captures the generating process---can forecast the
future behavior of the observed system with high accuracy.  [[Liz
    question: if there's noise, is that true?]]

Unfortunately, as a corollary of the undecidability of the halting
problem \cite{halting-problem}, there cannot exist \emph{one}
[[universally]] [[across {\sl all} time series?  this seems pretty
    obvious...]] ideal forecasting scheme for even completely
noise-free deterministic time series\cite{weigend-book}---let alone an
arbitrary real-world time series.  This naturally leads to an
important and challenging question: given a complicated real-valued
time series, does there exist a forecast model that can utilize the
information (if any) that is being transmitted from past to future by
the underlying generating process?  A first step in answering this
question is to reliably quantify where on the complexity spectrum a
given time series falls; a second step is to determine how complexity
and predictability are related.  These are the goals of this paper.

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
predict.  An important practical corollary to this is assessment of
forecast methods: if the forecast produced by a particular method is
poor but the time series contains a significant amount of predictive
structure, the method is inadequate and one should seek another
method.

We focus on real-valued, scalar, time-series data that appear
\emph{complicated}: i.e., not periodic, linear, constant, etc.  We
make no assumptions about the properties of the generating process:
whether it is linear, nonlinear, deterministic, stochastic,
stationary, non-stationary, etc.  The information in an observation
can be partitioned into two pieces: redundancy and entropy
generation\cite{crutchfield2003}.  Our approach exploits this
decomposition in order to assess how much predictive structure is
present in a signal---i.e., where it falls on the complexity spectrum
mentioned above.  We define \emph{complexity} as a particular
approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.  That is,
we view a random-walk time series (which exhibits high entropy) as
purely complex, whereas a periodic signal, which exhibits low entropy,
is on the low end of the complexity spectrum.  We argue that an
extension of \emph{permutation entropy} \cite{bandt2002per}---a method
for approximating the entropy through ordinal analysis---is an
effective way to assess the complexity of a given time series.
Permutation entropy is ideal for our purposes because it is known to
converge to the true entropy value. Other existing techniques either require specific knowledge of the generating process orproduce biased values of the entropy~\cite{bollt2001}.

To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a number of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is
  quantifiable by permutation entropy and correlated with prediction
  accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema, and in the choice of which forcasting scheme is appropriate.
\end{enumerate}
The forecast methods used here were chosen to span the space of
standard prediction strategies, but they do not cover that space
exhaustively.  Our goal here is an empirical assessment of the
relationship between predictability and complexity, not formal results
about a ``best'' predictor for a given time series.  From a
practitioner's standpoint, it would be useful to know {\sl a priori}
if a trace contains enough predictive structure to make it worth
spending the time and effort searching for a good forecast method.  It
would also be useful to know if a given method is inadequate---that
is, if one could do better and should keep searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  Computers are among the most complex
engineered artifacts in current use.  Modern microprocessor chips
contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
move data and threads of computation across those resources.  As a
result, the processor and memory loads during the execution of even a
very simple program can evolve in a very complicated fashion.
Figure~\ref{fig:ipc} shows a performance trace of a three-line loop
that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Radical changes in this behavior can result if the
program is changed slightly.  The resulting traces span the whole
range of the complexity spectrum, from completely predictable to
completely unstructured, making this an ideal testbed for this study.

@@

[[Following belongs in the methods section:]] In addition, some
performance traces even exhibit interesting regime changes as a
program switches between varying subroutines.

[[Following belongs in the conclusion.  Putting it here would be
    appropriate in a CS paper, but not in a physics journal, and not
    when we've just made the point that this is just a testbed.]] 
% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU, e.g., processor efficiency and similar statistics. Prediction
of these quantities at some finite time in the future, even with
perfect accuracy, does not violate the Rice-Shapiro
theorem~\cite{hopcroft2007} which essentially states that any non-trivial
property of a program is uncomputable.

>>>>

Complicated time-series data are ubiquitous in modern scientific
research.  The complexity of these data spans a wide range.  On the
low end of this spectrum are time series that exhibit perfect
predictive structure, i.e, signals whose future values can be
successfully predicted from past values.  In situations like this,
there is an underlying process that generates information and/or
transmits it from the past to the future in a perfectly predictable
fashion.  Constant or periodic signals, for example, fall in this
class.  On the opposite end of this spectrum are signals that are what
we call \emph{fully complex}, where the underlying generating process
transmits no information at all from the past to the future.  White
noise and random-walk processes fall in this class.  In fully complex
signals, knowledge of the past gives no insight into the future,
\emph{regardless} of what model one chooses to use.  Signals in the
midrange of this spectrum---e.g., deterministic chaos---pose
interesting challenges from a modeling perspective.  In these signals,
enough information is being transmitted from the past to the future
that an \emph{ideal} model---one that captures the generating
process---can forecast the future behavior of the observed system with
high accuracy.

As a corollary of the undecidability of the halting problem
\cite{halting-problem}, however, no single forecasting schema is ideal
for all noise-free deterministic signals\cite{weigend-book}---let
alone all real-world time series data sets.  This leads naturally to
an important and challenging question: given a complicated real-valued
time series, does there exist a forecast model that can utilize the
information (if any) that is being transmitted from past to future by
the underlying generating process?  A first step in answering this
question is to reliably quantify where on the complexity spectrum a
given time series falls; a second step is to determine how complexity
and predictability are related.  These are the goals of this paper.

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
predict.  An important practical corollary to this is a strategy for
assessing appropriateness of forecast methods.  If the forecast
produced by a particular method is poor but the time series contains a
significant amount of predictive structure, that method is inadequate
to the task and one should seek another method.  We focus on
real-valued, scalar, time-series data that appear complicated: i.e.,
not periodic, linear, constant, etc.  We make no assumptions about the
properties of the generating process: whether it is linear, nonlinear,
deterministic, stochastic, stationary, non-stationary, etc.

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation\cite{crutchfield2003}.  Our approach
exploits this decomposition in order to assess how much predictive
structure is present in a signal---i.e., where it falls on the
complexity spectrum mentioned above.  We define \emph{complexity} as a
particular approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.
That is, we view a random-walk time series (which exhibits high
entropy) as purely complex, whereas a periodic signal, which exhibits
low entropy, is on the low end of the complexity spectrum.  We argue
that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy through
ordinal analysis---is an effective way to assess the complexity of a
given time series.  Permutation entropy is ideal for our purposes
because it is known to converge to the true entropy value. Other
existing techniques either require specific knowledge of the
generating process or produce biased values of the
entropy~\cite{bollt2001}.

To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a number of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is
  quantifiable by permutation entropy and correlated with prediction
  accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which forecasting scheme is appropriate.
\end{enumerate}
The forecast methods used here were chosen to span the space of
standard prediction strategies, but they do not cover that space
exhaustively.  Our goal here is an empirical assessment of the
relationship between predictability and complexity, not formal results
about a ``best'' predictor for a given time series.  From a
practitioner's standpoint, it would be useful to know {\sl a priori}
if a trace contains enough predictive structure to make it worth
spending the time and effort searching for a good forecast method.  It
would also be useful to know if a given method is inadequate---that
is, if one could do better and should keep searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  Computers are among the most complex
engineered artifacts in current use.  Modern microprocessor chips
contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
move data and threads of computation across those resources.  As a
result, the processor and memory loads during the execution of even a
very simple program can evolve in a very complicated fashion.
Figure~\ref{fig:ipc} shows a performance trace of a three-line loop
that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the program can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting its \emph{performance}, which does not violate the
  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

@@
